---
title: "Impact of Severe Weather in the US"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Impact of Severe Weather in the US

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.


## Synopsis

The purpose of the assignment is to explore the NOAA Storm Database and analyse the effects of severe weather events on both population and economy.
This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The database covers the time period between 1950 and November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

The analysis aims to investigate which different types of sever weather events are most harmful on the populations health in respect of general injuries and fatalities. Further the economic consequences will be analyzed by exploring the financial damage done to both general property and agriculture (i.e. crops)

## Data Processing

### Data Loading:
Load and read the raw data file.

```{r echo=TRUE}
data <- read.csv("repdata_data_StormData.csv")
```

Explore what's contained in the dataset.

```{r echo=TRUE}
str(data)
```

### Data Subsetting

Subset the dataset on the parameters of interest.

```{r echo=TRUE}

    # Change parameter names to lowercase.
    colnames(data) <- tolower(colnames(data))

    # Subset on the parameters of interest.
    data <- subset(x=data, 
                   subset=(evtype != "?" & 
                               (injuries > 0 | fatalities > 0 | propdmg > 0 | cropdmg > 0)),
                   select=c("evtype", 
                            "fatalities", 
                            "injuries", 
                            "propdmg", 
                            "propdmgexp", 
                            "cropdmg", 
                            "cropdmgexp"))  
```

### Data Cleansing

Map the property and crop damage exponent alphabetic multipliers to numeric values.

```{r echo=TRUE}
    # Change all damage exponents to uppercase.
    data$propdmgexp <- toupper(data$propdmgexp)
    data$cropdmgexp <- toupper(data$cropdmgexp)

    # Map property damage alphanumeric exponents to numeric values.
    propDmgKey <-  c("\"\"" = 10^0,
                     "-" = 10^0, 
                     "+" = 10^0,
                     "0" = 10^0,
                     "1" = 10^1,
                     "2" = 10^2,
                     "3" = 10^3,
                     "4" = 10^4,
                     "5" = 10^5,
                     "6" = 10^6,
                     "7" = 10^7,
                     "8" = 10^8,
                     "9" = 10^9,
                     "H" = 10^2,
                     "K" = 10^3,
                     "M" = 10^6,
                     "B" = 10^9)
    data$propdmgexp <- propDmgKey[as.character(data$propdmgexp)]
    data$propdmgexp[is.na(data$propdmgexp)] <- 10^0
    
    # Map crop damage alphanumeric exponents to numeric values
    cropDmgKey <-  c("\"\"" = 10^0,
                     "?" = 10^0, 
                     "0" = 10^0,
                     "K" = 10^3,
                     "M" = 10^6,
                     "B" = 10^9)
    data$cropdmgexp <- cropDmgKey[as.character(data$cropdmgexp)]
    data$cropdmgexp[is.na(data$cropdmgexp)] <- 10^0
```

### Human Heath Data Processing

Select the applicable health columns from the dataset, then calculate the total number of fatalities and injuries per event type.

```{r echo=TRUE}
    # Aggregate number of fatalities and injuries per evtype into healthData dataframe
    healthData <- aggregate(cbind(fatalities, injuries) ~ evtype, data = data, FUN = sum)
    # Add total column to healthData
    healthData$total <- healthData$fatalities + healthData$injuries
```

Find the event types corresponding with the the highest health impacts.

```{r echo=TRUE}
    # Remove rows with zero health impact
    healthData <- healthData[healthData$total > 0, ]
    # Sort health data in descending order
    healthData <- healthData[order(healthData$total, decreasing=TRUE), ]
    # Re-label the rows
    rownames(healthData) <- 1:nrow(healthData)
    # Create dataframe of highest health impacting event types and append an "other" event type as a catchall 
    # for everything else
    healthDataTop <- healthData[1:10, ]
```   

### Economic Data Processing

Combine the damage and damage exponent multiplier parameters into the single parameters propertyloss and croploss.

```{r echo=TRUE}
    # Combine propdmg and propdmgexp parameters into a single parameter called propertyloss.
    data$propertyloss <- data$propdmg * data$propdmgexp
    # Combine cropdmg and cropdmgexp parameters into a single parameter called croploss.
    data$croploss <- data$cropdmg * data$cropdmgexp
```   


Select the applicable economic columns from the dataset, then calculate the total amount of property loss and crop loss per event type.

```{r echo=TRUE}
    # Aggregate amount of proploss and croploss per evtype into economicData dataframe
    economicData <- aggregate(cbind(propertyloss, croploss) ~ evtype, data=data, FUN=sum)
    # Add total loss column to economicData
    economicData$total <- economicData$propertyloss + economicData$croploss
```   
    
Find the event types corresponding with the highest economic impacts.

```{r echo=TRUE}
    # Remove rows with zero economic impact
    economicData <- economicData[economicData$total > 0, ]
    # Sort the economy data in descending order
    economicData <- economicData[order(economicData$total, decreasing=TRUE), ]
    # Re-label the rows
    rownames(economicData) <- tolower(rownames(economicData))
    # Create dataframe of highest economy impacting event types
    economicDataTop <- economicData[1:10, ]
```

## Results

### Figure 1: Top 10 Health Impact Event Types
Plot of the ten event types with the highest fatality counts plus an eleventh catchall event type that combines the total fatality counts of all other event types.

```{r echo=TRUE}
    # Load necessary libraries
    library(reshape2)
    library(ggplot2)

    # Melt the data
    healthDataTopMelt <- melt(healthDataTop, id.vars="evtype")

    # Create chart
    healthChart <- ggplot(healthDataTopMelt, aes(x=reorder(evtype, -value), y=value))
    # Plot data as bar chart
    healthChart = healthChart + geom_bar(stat="identity", aes(fill=variable), position="dodge")
    # Format y-axis scale and set y-axis label
    healthChart = healthChart + scale_y_sqrt("Frequency Count") 
    # Set x-axis label
    healthChart = healthChart + xlab("Event Type") 
    # Rotate x-axis tick labels 
    healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))
    # Set chart title
    healthChart = healthChart + ggtitle("Pareto Chart of Top 10 US Storm Health Impacts")
    # Display the chart
    print(healthChart)
``` 

### Figure 2: Top 10 Economic Impact Event Types
Plot of the ten event types with the highest economic impacts.

```{r echo=TRUE}
    # Load necessary libraries
    library(reshape2)
    library(ggplot2)

    # Melt the data
    economicDataTopMelt <- melt(economicDataTop, id.vars="evtype")

    # Create chart
    economicChart <- ggplot(economicDataTopMelt, aes(x=reorder(evtype, -value), y=value))
    # Add bars                            
    economicChart <- economicChart + geom_bar(stat="identity", aes(fill=variable), position="dodge")
    # Format y-axis scale and set y-axis label
    economicChart <- economicChart + scale_y_sqrt("Damage Impact [$]") 
    # Set x-axis label
    economicChart <- economicChart + xlab("Event Type") 
    # Rotate x-axis tick labels 
    economicChart <- economicChart + theme(axis.text.x = element_text(angle=45, hjust=1))
    # Set chart title
    economicChart <- economicChart + ggtitle("Pareto Chart of Top 10 US Storm Economic Impacts")
    # Display the chart
    print(economicChart)
```

## Conclusions

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Tornadoes are responsible for the largest proportion of both deaths and injuries out of all event types. 

### Across the United States, which types of events have the greatest economic consequences?

Flooding is responsible for the largeset proportion of total economic impact out of all event types.  
